# The psych package {#sec-psych}

## Introduction

In this chapter, I will focus exclusively on the `psych` package. I will cover all of the functions that I regularly use.

### Prerequisites

If you do not already have the `psych` package downloaded, you will first need to run `install.packages("psych")`

Then, you will be able to load the library and use its functions.

```{r}
#| label: setup
#| message: false

library(psych)
```

We will be using the `bfi` dataset that is included in the `psych` package. It is described as follows: *25 personality self report items taken from the International Personality Item Pool (ipip.ori.org) were included as part of the Synthetic Aperture Personality Assessment (SAPA) web based personality assessment project. The data from 2800 subjects are included here as a demonstration set for scale construction, factor analysis, and Item Response Theory analysis. Three additional demographic variables (sex, education, and age) are also included.*

```{r}
#| label: data
#| message: false
data(bfi)
```

## Describing Data {#sec-describe}

The `describe()` function will return the following for any numeric variable: number of valid cases, mean, standard deviation, trimmed mean (with trim defaulting to .1) , median (standard or interpolated, mad: median absolute deviation (from the median), minimum, maximum, skew, kurtosis, standard error.

Below I am `describing` the `A2` variable 
which has responses to the Agreeableness statement: `Inquire about others' well-being`. 

```{r}
describe(bfi$A2)
```

You can also describe by `groups` which will return descriptive statistics by groups. Below I am describing the `A2` variable by gender. Males = 1, Females =2.  

```{r}
describeBy(A2 ~ gender, data = bfi)
```

Another useful function is `headTail` which will show the first and last couple of lines of a dataset.

```{r}
headTail(bfi)
```


## Scoring Scales
